lifespan_eda

### 2025-01-25 eda/lifespan_eda.Rmd
filter phenos < or > 20 MADs from median

Load data from Jakob - updated with euler numbers and corrected VETSA study
```{r}
cn.df <- fread(paste0(base_path,"/data/v1_2_1_nature.csv"), stringsAsFactors = TRUE, na.strings="")
```

@@

left_join(cn.df.eulerfilt, study_mad) %>%
  ggplot() +
  geom_boxplot(aes(x=study, y=surfholes)) +
  geom_point(aes(x=study, y=(study_euler_median+2*study_euler_mad)), color="red")

@@

```{r}
#drop NAs listwise
cn.df.imp <- cn.df.imp1%>%
  dplyr::select(!c(INDEX.ID)) %>%#drop index ID for filtering
  na.omit()

@@

@@

```{r}
#sample size, sex
length(unique(cn.df.imp$participant))
nrow(cn.df.imp) == length(unique(cn.df.imp$participant))

table(cn.df.imp$sex)

table(cn.df.imp$sex)["Female"]/nrow(cn.df.imp)

#studies, sites
length(unique(cn.df.imp$study))
length(unique(cn.df.imp$site))

#age
min(cn.df.imp$age_days)/365.25-(280/365.25)
max(cn.df.imp$age_days)/365.25-(280/365.25)
mean(cn.df.imp$age_days)/365.25-(280/365.25)

abcd <- cn.df.imp %>%
  dplyr::filter(study=="ABCD")
length(unique(abcd$site))
```

```{r}
ggplot(cn.df.imp) +
  geom_violin(aes(x=reorder(study, age_days), y=age_days), scale="width") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggplot(cn.df.imp) +
  geom_violin(aes(x=reorder(site, age_days), y=age_days, fill=study), scale="width") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position = "none")
```

@@

```{r}
fwrite(cn.df.imp, file=paste0(base_path, "/data/lifespan_CN_imp-sites_euler.csv"))

#save names with site (since not saved correctly from centile outputs)
study_site_key <- cn.df.imp %>%
  dplyr::select(study, site, country, fs_version) %>%
  unique()
n_unique(study_site_key$study) == n_unique(cn.df.imp$study)
n_unique(study_site_key$site) == n_unique(cn.df.imp$site)
fwrite(study_site_key, file=paste0(base_path, "/data/lifespan_study_site_key.csv"))
```

@@

Clean up names, order by age
```{r}
#reorder studies by age
age.df <- cn.df.imp %>%
  group_by(study) %>%
  summarize(mean_age = mean(age_days)) %>%
  arrange(mean_age)
study_names_clean <- gsub("_", " ", unique(age.df$study))

final_df <- cn.df.imp %>% 
  mutate(age_years=(age_days-280)/365.25,
         study = gsub("_", " ", study)) %>%
  mutate(study = factor(study, levels=study_names_clean, ordered=TRUE))
```

@@

```{r}
sample_boxplot2 <- final_df %>%
  ggplot(aes(y=age_years, x=study, fill=country)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  #guides(y = ggh4x::guide_axis_nested(delim = "&")) +
  theme(axis.title.x=element_blank(), axis.text.x = element_blank(), legend.position = "right") +
  labs(fill="Country", y = "Age (Years)")  +
  paletteer::scale_fill_paletteer_d("ggthemes::Tableau_20")
```

@@

#overall sample
full_boxplot2 <- final_df %>%
  mutate(xlab = "Full Sample ") %>%
  ggplot(aes(x=xlab, y=age_years, color=sex)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 11) +
  theme(axis.title=element_blank(), axis.text=element_blank(), axis.ticks.y=element_blank()) +
  labs(color="Sex", x=" ")

sample_barplot2 <- final_df %>%
  ggplot() +
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], x=study, fill=sex)) +
  geom_hline(yintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_text(angle=90, hjust=0.5), axis.text.y = element_text(size=6)) +
  labs(fill="Sex", x = "Study", y="Percent Sex")

full_barplot2 <- 
  final_df %>%
  mutate(xlab = "  Full Sample  ") %>%
  ggplot() +
  geom_bar(aes(y=..count../tapply(..count.., ..x.. ,sum)[..x..], x=xlab, fill=sex)) + 
  geom_hline(yintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 11) +
  theme(axis.ticks.y = element_blank(), axis.title=element_blank(), axis.text.y=element_blank(), axis.text.x = element_text(angle=90, hjust=0.5, size=12)) +
  labs(color="Sex", y = "Age (Years)", y="")

full_rain <- final_df %>%
  ggplot(aes(x=sex, y=age_years, fill=sex)) +
  geom_rain(alpha=.4) +
  theme_linedraw(base_size = 11) +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_text(angle=90, hjust=0.5), 
        axis.title.x = element_blank(), axis.text.y = element_text(size=6),
        legend.position = "none") +
  labs(y="Age (Years)")
```

@@

fig5A <- ggarrange(sample_boxplot2, country_leg,
          sample_barplot2,  sex_leg,
          nrow=2, ncol=2, 
          #align = "v", 
          common.legend = FALSE,
          legend = "none",
          widths=c(11, 1),
          heights=c(1.9, 1)
          )
```

```{r}
fig5_AB <- ggarrange(fig5A, full_rain,
          nrow=1, ncol=2, 
          legend = "none",
          widths=c(3, 1),
          labels = "AUTO")

saveRDS(fig5_AB, file="../figs/figure_5_AB.rds")
```

>>>>

Load data from Jakob - updated with euler numbers and corrected VETSA study

```{r}
cn.df <- fread(paste0(base_path,"/data/v1_2_1_nature.csv"), stringsAsFactors = TRUE, na.strings="")
pheno_list <- readRDS(paste0(base_path, "/cubic_scripts/R_scripts/pheno_list.rds"))
```

@@

left_join(cn.df.eulerfilt, study_mad) %>%
  ggplot() +
  geom_boxplot(aes(x=study, y=surfholes)) +
  geom_point(aes(x=study, y=(study_euler_median+2*study_euler_mad)), color="red") +
  theme(axis.text.x = element_text(angle=90, hjust=0.5))

@@

```{r}
#make sure index.id is ok for filtering
stopifnot(sum(is.na(cn.df.imp1$INDEX.ID))==0)

#drop NAs listwise
cn.df.imp <- cn.df.imp1%>%
  na.omit()

@@

Finally, drop 0 phenos or biologically implausible

```{r}
#find 0 IDPs
df_zeros <- cn.df.imp %>%
  filter(if_any(any_of(pheno_list), ~ . == 0 & !is.na(.))) %>%
  filter(if_any(all_of(pheno_list), ~ !is.na(.)))
nrow(df_zeros)
```

no one with IDPs=0

check for implausible values

```{r}
#check there aren't currently any NAs
stopifnot(sum(is.na(cn.df.imp))==0)

cn.df_mad <- lapply(cn.df.imp[c("INDEX.ID", pheno_list)], mad_filt, n=20) %>% as.data.frame()
sum(is.na(cn.df_mad))

mad_drop <- filter(cn.df_mad, if_any(pheno_list, is.na))

#remove 3 subjects
cn.df_mad_filt <- cn.df.imp %>%
  dplyr::filter(! INDEX.ID %in% mad_drop$INDEX.ID)

nrow(cn.df.imp) == (nrow(cn.df_mad_filt) + nrow(mad_drop))
```

@@

```{r}
#sample size, sex
length(unique(cn.df_mad_filt$participant))
nrow(cn.df_mad_filt) == length(unique(cn.df_mad_filt$participant))

table(cn.df_mad_filt$sex)

table(cn.df_mad_filt$sex)["Female"]/nrow(cn.df_mad_filt)

#studies, sites
length(unique(cn.df_mad_filt$study))
length(unique(cn.df_mad_filt$site))

#age
min(cn.df_mad_filt$age_days)/365.25-(280/365.25)
max(cn.df_mad_filt$age_days)/365.25-(280/365.25)
mean(cn.df_mad_filt$age_days)/365.25-(280/365.25)

abcd <- cn.df_mad_filt %>%
  dplyr::filter(study=="ABCD")
length(unique(abcd$site))
```

@@

```{r}
fwrite(cn.df_mad_filt, file=paste0(base_path, "/data/lifespan_CN_imp-sites_euler.csv"))

#save names with site (since not saved correctly from centile outputs)
study_site_key <- cn.df_mad_filt %>%
  dplyr::select(study, site, country, fs_version) %>%
  unique()
n_unique(study_site_key$study) == n_unique(cn.df_mad_filt$study)
n_unique(study_site_key$site) == n_unique(cn.df_mad_filt$site)
fwrite(study_site_key, file=paste0(base_path, "/data/lifespan_study_site_key.csv"))
```

@@

Clean up names, order by age
```{r}
#reorder studies by age
age.df <- cn.df_mad_filt %>%
  group_by(study) %>%
  summarize(mean_age = mean(age_days)) %>%
  arrange(mean_age)
study_names_clean <- gsub("_", " ", unique(age.df$study))

final_df <- cn.df_mad_filt %>% 
  mutate(age_years=(age_days-280)/365.25,
         study = gsub("_", " ", study)) %>%
  mutate(study = factor(study, levels=study_names_clean, ordered=TRUE))
```

@@

```{r}
sample_boxplot2 <- final_df %>%
  ggplot(aes(y=age_years, x=study, fill=country)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 14) +
  #guides(y = ggh4x::guide_axis_nested(delim = "&")) +
  theme(axis.title.x=element_blank(), axis.text.x = element_blank(), legend.position = "right") +
  labs(fill="Country", y = "Age (Years)")  +
  paletteer::scale_fill_paletteer_d("ggthemes::Tableau_20")
```

@@

#overall sample
full_boxplot2 <- final_df %>%
  mutate(xlab = "Full Sample ") %>%
  ggplot(aes(x=xlab, y=age_years, color=sex)) +
  geom_boxplot(outlier.size = 1, outlier.colour=NULL) +
  theme_linedraw(base_size = 14) +
  theme(axis.title=element_blank(), axis.text=element_blank(), axis.ticks.y=element_blank()) +
  labs(color="Sex", x=" ")

sample_barplot2 <- final_df %>%
  ggplot() +
  geom_bar(aes( y=..count../tapply(..count.., ..x.. ,sum)[..x..], x=study, fill=sex)) +
  geom_hline(yintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 14) +
  scale_y_continuous(labels = scales::percent) +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_text(angle=90, hjust=0.5), axis.text.y = element_text(size=10)) +
  labs(fill="Sex", x = "Study", y="Percent Sex")

full_barplot2 <- 
  final_df %>%
  mutate(xlab = "  Full Sample  ") %>%
  ggplot() +
  geom_bar(aes(y=..count../tapply(..count.., ..x.. ,sum)[..x..], x=xlab, fill=sex)) + 
  geom_hline(yintercept = .5, color="lightgray", linetype="longdash") +
  theme_linedraw(base_size = 14) +
  theme(axis.ticks.y = element_blank(), axis.title=element_blank(), axis.text.y=element_blank(), axis.text.x = element_text(angle=90, hjust=0.5)) +
  labs(color="Sex", y = "Age (Years)", y="")

```

```{r}
full_rain <- final_df %>%
  mutate(continent = factor(case_when(country == "India" | country == "China" | country == "Singapore" | country == "Japan" ~ "Asia",
                               country == "USA" | country == "Canada" ~ "North America",
                               country == "Brazil" ~ "South America",
                               country == "Australia" ~ "Australia",
                               TRUE ~ "Europe"))) %>%
  ggplot(aes(x=continent, y=age_years, fill=sex)) +
  geom_rain(alpha=.5, violin.args = list(scale="count", alpha=.5),
            point.args=list(alpha=.2)) +
  theme_linedraw(base_size = 14) +
  theme(axis.title.x = element_blank(), axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "bottom") +
  guides(fill=guide_legend(title="Sex")) + 
  coord_fixed(ratio=.08) +
  labs(y="Age (Years)")
```

@@

fig5A <- ggarrange(sample_boxplot2, country_leg,
          sample_barplot2,  sex_leg,
          nrow=2, ncol=2, 
          #align = "v", 
          common.legend = FALSE,
          legend = "none",
          widths=c(8, 1),
          heights=c(1.9, 1)
          )
```

```{r}
fig5_AB <- ggarrange(fig5A, full_rain,
          nrow=1, ncol=2, 
          #legend = "none",
          widths=c(1.8, 1),
          labels = "AUTO")
fig5_AB
saveRDS(fig5_AB, file="../figs/figure_5_AB.rds")
```
